HO unification from object language to meta language

### 2024-04-01 paper/main.tex
more \def, adjust printer_pb (links on 3 col + empty T)

\usepackage{myTools}
\usepackage{macros}
\usepackage{enumitem}

@@

\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}
\def\phantx{\phantom{x}}

@@

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
unification problem between terms $\mathbb{S}_{p_l}$ and
$\mathbb{S}_{p_r}$ taken from the set of all terms $\mathbb{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
%
$$
\begin{array}{l}
\stepF{\mathbb{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathbb{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathbb{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathbb{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathbb{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathbb{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathbb{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathbb{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathbb{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathbb{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation]\label{prop:simulation}
$\forall \mathbb{S}, \forall \mathcal{N},$
$$
  \runF{\mathbb{S}}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\mathbb{S}}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathbb{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\mathbb{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathbb{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathbb{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathbb{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\mathbb{T}) \land
  \sigma\mathbb{T}_{p_l} \Ue \sigma\mathbb{T}_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \mathbb{T})\\
\wellb(\sigma\mathbb{T}) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \mathbb{T})
\end{array}
$$
\end{proposition}

@@

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}
The basic compilation scheme is not about to
deal wit the following problem:
\printAlll
  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,G}}}
  {{}}
  {{}}
  {{}}
\marginpar{TODO: remove empty T}
% \begin{gather}
% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
% \end{gather}

noindent Note that here $X$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
Both problems are addressed in the next two sections.

@@

Following the compilation scheme of \cref{sec:compilation} the
unification problem $\mathbb{P}$ is compiled as follows:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

\noindent
While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
$\mathbb{T}$ does not:
\elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors hence
\Ue{} fails.

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
we modify the term compiler put fresh unification variables
in their place: the problematic term is moved from 
from  $\mathbb{T}$ to \linkStore (\cref{sec:etacomp}). The compilation
of the problem $\mathbb{P}$ above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or variable applied to arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta{} T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\marginpar{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect\cref{inv:map-bijective} (merging the two mappings
for $s$ would break \ref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

>>>>

\usepackage{myTools}
\usepackage{macros}
\usepackage{enumitem}
\usepackage{intcalc}

@@

\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}

\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
\newcommand{\foUnifPb}{\ensuremath{\mathbb{P}}\xspace}
\newcommand{\hoUnifPb}{\ensuremath{\mathbb{T}}\xspace}

@@

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
unification problem between terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation]\label{prop:simulation}
$\forall \foUnifPb, \forall \mathcal{N},$
$$
  \runF{\foUnifPb}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\foUnifPb}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\hoUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\foUnifPb = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\hoUnifPb) \land
  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
\wellb(\sigma\hoUnifPb) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \hoUnifPb)
\end{array}
$$
\end{proposition}

@@

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}
The basic compilation scheme is not about to
deal wit the following problem:
\printAlll
  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
  {{}}
  {{}}
  {{}}
% \begin{gather}
% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
% \end{gather}

\noindent Note that here $X$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
Both problems are addressed in the next two sections.

@@

Following the compilation scheme of \cref{sec:compilation} the
unification problem \foUnifPb is compiled as follows:
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

\noindent
While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
\hoUnifPb does not:
\elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors hence
\Ue{} fails.

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
we modify the term compiler by putting fresh unification variables
in their place: the problematic term is moved from 
from  \hoUnifPb to \linkStore (\cref{sec:etacomp}). The compilation
of the problem \foUnifPb above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or variable applied to arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\marginpar{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

We report here the problem given in \cref{sec:basic-comp-limitations} where
$X$ is used with two different arities and the output of the compilation does
not respect \cref{inv:map-bijective} (merging the two mappings
for $s$ would break \cref{inv:uvaarity}).
In this section we explain how to replace the duplicate mapping with some
\linketa{} in order to restore the invariants.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}
